---
title: '\ '
output: blogdown::html_page
---

```{r setup, echo = F, message = F, warning = F}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               here,
               knitr)

source(here("global/functions/misc_functions.R"))

opts_chunk$set(warning = F, message = F)
```

## Prologue

This quiz tests your understanding of the `filter()` function in the **dplyr** package. Good luck!

For multiple choice questions, please enter the correct letter option into the provided box.

------------------------------------------------------------------------

## Question-1-1

<!--# Objective: Practice reading "filter( CONDITION )" as "keep rows matching CONDITION", and filter(! CONDITION ) as "drop rows matching CONDITION". -->

Which of the following is a **misinterpretation** of a **filter** statement on `df`? Answer: {a}

a.  `df %>% filter(age >  5)` : Drop row(s) where `age` is above 5

b.  `df %>% filter(age <  5)` : Keep row(s) where `age` is below 5

c.  `df %>% filter(!age ==  5)` : Drop row(s) where `age` is equal to 5

------------------------------------------------------------------------

## Question-1-2

Which of the following is a **misinterpretation** of a **filter** statement on `df`? Answer: {a}

a.  `df %>% filter(deaths <  4)` : Drop row(s) where `deaths` is below 4

b.  `df %>% filter(!deaths >  4)` : Drop row(s) where `deaths` is above 4

c.  `df %>% filter(deaths ==  4)` : Keep row(s) where `deaths` is equal to 4

------------------------------------------------------------------------

## Question-1-3

Which of the following is a **misinterpretation** of a **filter** statement on `df`? Answer: {b}

a.  `df %>% filter(cases !=  3)` : Keep row(s) where `cases` is not equal to 3

b.  `df %>% filter(!cases >  3)` : Keep row(s) where `cases` is above 3

c.  `df %>% filter(cases ==  3)` : Keep row(s) where `cases` is equal to 3

------------------------------------------------------------------------

## Question-2-1

<!--# Objective: Filter on NA-containing columns. -->

`df` is an excerpt of a edataset from an H7N9 outbreak in China.

```{r render = head_5_rows}
df <- outbreaks::fluH7N9_china_2013[c(1,2,3,74, 75), c(6:8)]
df
```

To **drop** the single female patient, which of the following **filter** statements is **correct** Answer: {a}

a.  `df %>% filter(gender ==  "m" & is.na(gender))`

b.  `df %>% filter(gender !=  "f")`

c.  `df %>% filter(gender ==  "m")`

------------------------------------------------------------------------

## Question-3-1

<!--# Objective: Use the `%in%` operator properly -->

```{r echo = F}
df <- 
  tribble(~id, ~state, 
           1,  "Kano",
           2,  "Lagos",
           3,  "Bauchi", 
           4,  "Kano", 
           5,  "FCT",
           6,  NA)

df
```

From the dataset shown, you would like to keep rows where `state` is either "Lagos" or "Kano". Which of the following uses **correct syntax** for this. Answer: {b}

a.  `df %>% filter(state %in% c("Lagos" "Kano"))`

b.  `df %>% filter(state %in% c("Lagos", "Kano"))`

c.  `df %>% filter(state == c("Lagos", "Kano"))`

------------------------------------------------------------------------

## Question-3-2

```{r echo = F}
df <- 
  tribble(~id, ~state, 
           1,  "Kano",
           2,  "Lagos",
           3,  "Bauchi", 
           4,  "Kano", 
           5,  "FCT",
           6,  NA)

df
```

From the dataset shown, you would like to keep rows where `state` is either "Bauchi" or "Kano". Which of the following uses **correct syntax** for this. Answer: {c}

a.  `df %>% filter(state %in% ("Bauchi" "Kano")`

b.  `df %>% filter(state %in% c("Bauchi" OR "Kano"))`

c.  `df %>% filter(state %in% c("Bauchi", "Kano"))`

------------------------------------------------------------------------

## Question-3-3

```{r echo = F}
df <- 
  tribble(~id, ~state, 
           1,  "Kano",
           2,  "Lagos",
           3,  "Bauchi", 
           4,  "Kano", 
           5,  "FCT",
           6,  NA)

df
```

From the dataset shown, you would like to keep rows where `state` is either "FCT" or "Kano". Which of the following uses **correct syntax** for this. Answer: {a}

a.  `df %>% filter(state %in% c("FCT", "Kano"))`

b.  `df %>% filter(state == "FCT" & state == "Kano"))`

c.  `df %>% filter(state == c("FCT", "Kano"))`

------------------------------------------------------------------------

## Question-4-1

Answer T for TRUE or F for FALSE.

In R, `=` can be used as an assignment operator, while `==` is used for testing equality. Answer: {[T][TRUE]}

------------------------------------------------------------------------

## Epilogue

If you struggled with this quiz, please consult the following resources:
