---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
#  word_document:
#    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
---

```{r, echo = FALSE, include = FALSE, warning = FALSE, message = FALSE}
# Load packages ----
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               here)

# Source functions ----
source(here::here("global/functions/misc_functions.R"))

# Set default render for dataframes ----
registerS3method("reactable_5_rows", "data.frame", reactable_5_rows)

# Set default chunk options ----
knitr::opts_chunk$set(class.source = "tgc-code-block", error = TRUE,
                      render = reactable_5_rows)

# Set template labels for custom chunk options ----
knitr::opts_template$set(
  example = list(echo = TRUE, eval = TRUE),
  practice = list(echo = FALSE, eval = FALSE),
  image1 = list(echo = FALSE, out.width = "75%", fig.align = "center"),
  image2 = list(echo=FALSE, out.width="75%", fig.align="center"),
  justcode = list(echo = TRUE, eval = FALSE)
)
```

# Learning objectives

1.  <objective>

2.  <objective>

3.  <objective>

4.  <objective>

5.  <objective>


# Examples for students to run
```{r <chunk_name>, opt.label = example}

```

# Images

## Method 1: print_image()
```{r <chunk_name>, opt.label = image1, fig.cap = "<caption>"}
print_image(here::here("CHAPTER_00_CHAPTER_NAME/images/FILE_NAME.png"))
```

## Method 2: include_graphics()
```{r <chunk_name>, opt.label = image2, fig.cap = "<caption>"}
knitr::include_graphics(path = "../images/FILE_NAME.EXT")
```

# Practice questions

## Two-question block

=======
# Learning objectives

1.  <objective>

2.  <objective>

3.  <objective>

4.  <objective>

5.  <objective>


# Examples for students to run
```{r <chunk_name>, opt.label = example}

```

# Images

## Method 1: print_image()
```{r <chunk_name>, opt.label = image1, fig.cap = "<caption>"}
print_image(here::here("CHAPTER_00_CHAPTER_NAME/images/FILE_NAME.png"))
```

## Method 2: include_graphics()
```{r <chunk_name>, opt.label = image2, fig.cap = "<caption>"}
knitr::include_graphics(path = "../images/FILE_NAME.EXT")
```

# Practice questions

## Two-question block

>>>>>>> 5fe0add7e35b3e53b99a5793c037111ed55c1651
::: practice
- <question_text>
```{r q_, opt.label = practice}
q_ <- "YOUR_ANSWER_HERE" 
.check_q_()
.hint_q_()
```

- <question_text>
```{r q_, opt.label = practice}
q_ <- "YOUR_ANSWER_HERE" 
.check_q_()
.hint_q_()
```
:::

## One-question block

::: practice
- <question_text>
```{r q_, opt.label = practice}
q_ <- "YOUR_ANSWER_HERE" 
.check_q_()
.hint_q_()
```
:::

# Display code
```{r <chunk_name>, opt.label = justcode}

```

# Build your own
```{r <chunk_name>, <chunk_options>}

```

# Score print
```{r grade_lesson, opt.label = practice}
.score_print()
```

# Other

## Key point block

::: key-point
-   <bullet point>
-   <bullet_point>
:::

## Watch out block

::: watch-out
-   <bullet_point>
-   <bullet_point>
:::

# Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("name"))`

# References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   <ref>

-   <ref>

`r tgc_license()`
