---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: console
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)

# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

# Source autograder script quietly 
mute(here("ch04_functions/lessons/ch04_ls01_functions_autograder.R"))
```

<!-- # Functions: introduction -->

## Learning objectives

1.  You will learn how to create your own functions.

2.  You will learn the importance of creating your own functions as you improve your career as a developing data scientist.

3.  You will learn how to write the body of a function.

4.  You will learn how to design and handle the arguments (parameters) of a function.

## Introduction to the dataset

In this lesson, we will continue to to use the COVID-19 dataset containing the results from a serological survey conducted in Yaounde, Cameroon in late 2020.

```{r render = reactable_10_rows, message = FALSE}
yaounde <- read_csv(here("ch04_functions/data/yaounde_data.csv"))
yaounde

```

## Introducing functions

R language is based mainly on using objects and functions. Quoting [John Chambers](https://statweb.stanford.edu/~jmc4/vitae.html), everything that "exists" in a R environment is an object, and everything that "happens" is a function.

So far, we have already encountered and used a large number of functions. Functions can be quite simple (such as `sum()`, `max()`, `mean()`), or really complex (such as what we are seeing in the `tidyverse` package like `mutate()`, and even more so functions related to statistical analysis which you will discover in further courses).

@ILLUSTRATE:

The basic principle behind a function is that it needs to receive as input one or more arguments (or parameters), to perform a certain number of actions so as to return a result. Functions are programmed in a way that the user does not see its code running; all calculations and actions are usually done in the background. Finally, the function will return the desired result, which can - but not always - be accompanied by a message or even an error. More on these later.

## Why should I bother writing functions?

Writing functions is a very important task to any data scientist, from beginners to seasoned coders. Functions not only make you more experienced with the R language, but it allows you to speed up your work and even reduce the chance of making errors while coding. You will be able to automate common and repetitive tasks in your data analysis flow (such as calculating epidemiological indicators, for example). This is very important because it will allow you to drop the dreaded copy-and-paste routine that plagues many good scripts (which can easily turn into hundreds or thousands of copied redundant lines). 

@ILLUSTRATE

The main advantages of writing functions over copy-and-pasting are:

- You can give a function an evocative name that makes your code easier to understand.

- As requirements change, you only need to update code in one place, instead of many.

- You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

## You convinced me, when do we start to writing our own functions?

In a few moments. But, before starting with our coding, you should know that writing your own functions is a lifetime journey. Even experienced programmers learn new techniques and better ways of approaching old problems. This happens because as we grow more experienced in our path as data scientists, we are exposed to new problems and solving them gives us new insights in turn. So, don't worry about mastering every little detail, because it is just not possible to list them all. This lesson will provide you with some advice that you can learn and start applying immediately.

Also, don't worry if things seem complicated. As said above, time and experience (and a lot of testing!) will make you write good and useful functions in no time!

::: practice
-   So far, what have you understood about writing your own functions in R?

    A. To be able to automate tasks that are needed in the script and minimize the chance of erring when we need to copy-and-paste chunks of code multiple times.

    B. Functions are a less effective way of repeating chunks of code that we can easily copy-and-paste in our script.

    C. It is easier to spot mistakes and typos when we are copying-and-pasting repeated chunks of code, since we are familiar with it.

```{r include = F}

q1 <- "A"

.check_q1()

```
:::

------------------------------------------------------------------------

## Basics of a function

Here is an example of a very simple function that only adds 2 to a number:

```{r}

sum_2 <- function(x) {
  result <- x + 2
  return(result)
}

```

If you execute this code, you will create a function named `sum_2`, which can be used directly in a script or in the console:

```{r}

sum_2(12)

```

Before going further, let's break down the structure of this first function step by step. First, a function is created using the statement `function`. This is followed by a pair of parentheses and a pair of braces.

```{r eval=FALSE}

function() {
  
  }

```

Inside the parentheses, we indicate the arguments of the function, those that must be passed to it when we call it. Our function only takes one argument, which we have arbitrarily decided to name `x`.

```{r eval=FALSE}

function(x) {
  
  }

```

The braces include a series of R statements that make up the **body** of the function. This is the code that will be executed when our function is called. The arguments passed to it must be used here inside the body of the function. In our function `sum_2`, the first line takes the value of the argument x, adds 2 to it, and stores the result in a new object `result`.

```{r eval=FALSE}

function(x) {
  result <- x + 2
}

```

Now we want that our function return what is calculated inside its body. This is achieved via the instruction `return` to which we pass the value to return (in our case, the object `result`).

```{r eval=FALSE}

function(x) {
  result <- x + 2
  return(result)
}

```

Finally, in order for our function to be called and used, we need to give it a name. This is the same as storing a value in an object. Here we store it in an object named `sum_2`.

```{r eval=FALSE}

sum_2 <- function(x) {
  result <- x + 2
  return(result)
}

```

With our function, we have therefore created a new object in our environment called `sum_2`, of type `function`. This new function takes a single argument `x`, calculates the value `x + 2` and returns that result. It is used by typing its name followed by the value of its argument in parentheses, in the same manner that we have done so far in our course. For example:

```{r}

sum_2(12)

```

Or:

```{r}

sum_2(x = 12)

```

You can also assign the value to an object and pass it to the function. This is what the pros do in their scripts, since it avoids what is called "hardcoding":

```{r}

value <- 12

sum_2(value)

```

::: pro-tip
**Why is hardcoding bad?**

Hardcoding is a common bad practice of inputing data directly into the code, as opposed to obtaining the data from an external source. Whenever possible, we should avoid hardcoding our scripts in order to prevent errors. In very long scripts, the act of hardcoding might cause unnexpected errors (especially if the user is a copy-and-paste addict). This is analogous to our initial examples of why you should write your own functions: if copying-and-pasting chunks of repeated code is by itself prone to errors, imagine copying-and-pasting chunks of code while manually changing some values (hardcoding). This is a review nightmare. In our example above, we avoided hardcoding by assigning the value to an object, which was then passed to the `sum_2` function.
:::

Note that if `x` is a vector, we can also call our function by passing `x` as a vector into the argument. Our vector will be a sequence of 10 up to 50, with jumps of 10 between (10, 20, 30, 40 and 50). This is achieved by the `seq` function in R base.

```{r}

vector <- seq(from = 10, to = 50, by = 10)

sum_2(vector)

```

This is specially important when working with datasets, since we can pull a variable out of the dataset to apply a function. Coming back to our dataset! Let's apply the function we have just created to the `age` variable of the Yaounde dataset (i.e. adding +2 to the age collected for each participant). We will do this only for the first 10 observations (notice the [1:10], used to select observations 1 through 10).

```{r}

age <- yaounde$age[1:10]

sum_2(age)

```

::: practice
1. Create a simple function called `age_months` that transforms age in years to age in months. (With the hint being: there are 12 months per year, think multiplication.)

```{r include = F}

age_months <- function(x) { 
                  months <- x * 12 
                  return(months) }

q2 <- age_months

.check_q2()

.hint_q2()

```
:::

::: practice
2. Now, apply your function to the yaounde dataset's age variable (i.e. convert the age in years of all participants to an age in months). You should submit a vector. 

```{r include = F}

years <- yaounde$age

q3 <-  q2(years)

.check_q3()

.hint_q3()

```
:::

```{r include = F}
.score_print()
```

## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("dcpcamara", "eduardo", "lolovanco", "kendavidn"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Barnier, Julien. "Introduction à R et au tidyverse." Accessed May 23, 2022. <https://juba.github.io/tidyverse>

-   Wickham, Hadley; Grolemund, Garrett. "R for Data Science." Accessed May 25, 2022. <https://r4ds.had.co.nz/>

<!-- (Chicago format. You can use https://www.citationmachine.net) -->

`r tgc_license()`
