---
title: '\ '
output: blogdown::html_page
params:
  seed: 1
---

```{r, echo = F, message = F, warning = F}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(
  # necessary for all quizzes runs
  tidyverse, 
  here,
  knitr,
  googledrive,
  httpuv,
  rio,
  # specific to this quiz             
  cowplot, 
  xfun)

knitr::opts_chunk$set(warning = F, message = F)

source(here("global/functions/misc_functions.R"))

# Setup for uploading images and datasets
if (!drive_has_token()) drive_auth(email = "trainingteam@thegraphnetwork.org")
options(gargle_oauth_email = "trainingteam@thegraphnetwork.org")
knitr::opts_knit$set(upload.fun = gdrive_upload) # pulled by knitr to upload image

# Set seed
set.seed(params$seed)
```

```{r echo = F}
ebola_dat_raw <- 
  outbreaks::ebola_sierraleone_2014 %>% 
  as_tibble() %>% 
  select(-chiefdom) %>% 
  filter(date_of_onset < as.Date("2014-06-30")) %>% 
  arrange(id) %>% 
  mutate(id = row_number()) %>% 
  # Sample data
  slice_sample(n = 200)
```

Here you will analyze a dataset sample of confirmed and suspected cases of Ebola hemorrhagic fever in Sierra Leone, in May and June of 2014.

```{r echo = F, results = "asis"}
upload_and_link(ebola_dat_raw, filename = "ebola_dat_raw", format =  "csv")
```

Here are the top 6 rows of `ebola_dat_raw` after import:

```{r}
head(ebola_dat_raw)
```

------------------------------------------------------------------------

# Part A

For some rows in the dataset, the "sex" variable is missing. We would like to drop such rows. Additionally, we are not interested in analyzing any cases from the Bo district.

**1. Complete the code to drop everyone whose sex is missing or who is from the Bo district**

```{r echo = F}
ebola_dat <- 
  ebola_dat_raw %>% 
  filter(!( is.na(sex) | district == "Bo" ))
```

```{r eval = F}
ebola_dat <- 
  ebola_dat_raw %>% 
  filter(!( {is.na}(sex) | {district} == "Bo" ))
```

```{r echo = F}
filter_drop_hint <- collapsible('Recall that `filter(!( CONDITIONS_HERE ))` can be read as "DROP every row matching any of the conditions surrounded by `!()`')
```

`r filter_drop_hint`

**2. How many rows remain in the filtered dataset? {`r nrow(ebola_dat)`}**

------------------------------------------------------------------------

```{r echo = F}
ebola_dat_drop_kids_and_na_age <- 
  ebola_dat_raw %>% 
  filter(!is.na(age) & !age < 18 )
```

**3. How many rows remain if you, instead, drop all children (\< 18) and anyone with a missing age from `ebola_dat_raw`. {`r nrow(ebola_dat_drop_kids_and_na_age)`}**

------------------------------------------------------------------------

# Part B

With the filtered dataset, `ebola_dat`, from question A1/A2, we would like to create a frequency table of cases by `district`.

**1. Complete the missing code that generates the frequency table whose "head" is shown. Then fill in the missing values in the output.**

```{r eval = F}
sex_district_count <- 
  ebola_dat %>%
  {count}({district})
sex_district_count
```

```{r include = F}
sex_district_count <- 
  ebola_dat %>%
  count(district)
sex_district_count
```

```{r echo = F}
sex_district_count %>% 
  mutate(across(.fns = as.character)) %>% 
  # Replace odd-numbered rows with fill-in-the-blank syntax
  mutate(n = case_when(row_number() %% 2 == 0 ~ paste0("{", n, "}"),
                       TRUE ~ n)) %>% 
  head() 
```

------------------------------------------------------------------------

Next, use a similar function to generate a frequency table of cases by `status` and `sex` for the `ebola_dat` data frame.

```{r echo = F}
status_sex_count <- 
  ebola_dat %>%
  count(status, sex)
```

**2. Based on this, how many suspected cases are there among women?** {`r status_sex_count %>% filter(status == "suspected", sex == "F") %>% .$n`}

**3. And how many confirmed cases are there among men?** {`r status_sex_count %>% filter(status == "confirmed", sex == "M") %>% .$n`}

------------------------------------------------------------------------

# Part C

From B1, we notice that the largest number of cases are in Kailahun. We now want to plot the case counts over time in this district.

**1. Complete the code that generates the plot shown.**

```{r eval = F}
ebola_dat %>% 
  filter(district == "Kailahun") %>% 
  ggplot(aes({[x=date_of_onset][date_of_onset]})) +
  geom_bar() + 
  scale_x_date(breaks = scales::pretty_breaks(n = 60)) + 
  {theme}(axis.text.x = element_text(angle= 90, vjust = 0.5))
```

```{r echo = F}
ebola_dat %>% 
  filter(district == "Kailahun") %>% 
  ggplot(aes(x = date_of_onset)) + 
  geom_bar() + 
  scale_x_date(breaks = scales::pretty_breaks(n = 60)) + 
  theme(axis.text.x = element_text(angle= 90, vjust = 0.5))
```

------------------------------------------------------------------------

**2. Now create a similar case-count plot for just *children* (\< 18) from *Kailahun*.**

```{r include = F}
children_kailahun <-
  ebola_dat %>%
  filter(age < 18, district == "Kailahun") %>% 
  select(date_of_onset)

children_kailahun %>% 
  ggplot(aes(x = date_of_onset)) + 
  geom_bar() + 
  scale_x_date(breaks = scales::pretty_breaks(n = 60)) + 
  theme(axis.text.x = element_text(angle= 90, vjust = 0.5))
```

```{r echo = F}
plotlist <- list()
correct_letter <- sample(letters[1:4], size = 1)

for (i in letters[1:4]) {
  
  # resample for wrong answers
  set.seed(match(i, letters)*2)
  
  children_kailahun_sampled <- if (i != correct_letter) {
    children_kailahun %>%
      slice_sample(n = 42, replace = TRUE)
  } else {
    children_kailahun
  }
  
  
  plotlist[[i]] <- 
    children_kailahun_sampled %>% 
    ggplot(aes(x = date_of_onset)) + 
    geom_bar() + 
    theme_void() + 
    theme(plot.background = element_rect(color = "black"))
    
}

plot_options <- plot_grid(plotlist$a, plotlist$b, plotlist$c, plotlist$d, 
                          labels = "auto")
```

**Which of the figures below represents this case-count plot for children from Kailahun? Letter answer : {`r correct_letter`}**

```{r echo = F}
plot_options
```

```{r include = F}
max_days_without_cases <- 
  children_kailahun %>% 
    count(date_of_onset) %>% 
    filter(date_of_onset >= as.Date("2014-06-01") & 
             date_of_onset <= as.Date("2014-06-30")) %>% 
    mutate(diffs = date_of_onset - lag(date_of_onset)- 1) %>% 
    .$diffs %>% 
    max(na.rm = T) %>% 
    as.numeric()
```

**3. In June, what was the longest number of days over which no cases were reported in Kailahun children? {[`r numbers_to_words(max_days_without_cases)`][`r max_days_without_cases`]} days**
