---
title: '\ '
output:
  html_document:
    number_sections: yes
    toc: yes
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r, echo = F, message = F, warning = F}
if(!require("pacman")) install.packages("pacman")
pacman::p_load("tidyverse",
               "knitr",
               "janitor",
               "stringr",
               "here",
               "skimr",
               "inspectdf",
               update=F)

skim_format(.levels = list(max_char = 4))
# Source functions 
source(here::here("global/functions/misc_functions.R"))

# default render
registerS3method("reactable_5_rows", c("data.frame","matrix"), reactable_5_rows)
knitr::opts_chunk$set(class.source = "tgc-code-block")

## autograders
suppressMessages(source(here::here(("ch02_data_cleaning_pipeline/lessons/ch02_ls05_transformation_autograder.R"))))

```

------------------------------------------------------------------------

# Data cleaning: data transformation

## Learning objectives

1.  You will learn how to fix inconsistencies and structural errors in the variables using `recode`, `replace`, and `gsub`
2.  You will verify datatypes
3.  You will correct datatypes using `across()` to transform multiple columns.

## Intro to the lesson

::: key-point
Data format determines the kind of manipulations and plotting possible with such data.

An important step in data cleaning is **checking the datatypes** of your variables and **fixing any inconsistencies** that would cost time and accuracy of results later on.
:::

::: pro-tip
A few elements you want to keep in mind as targets:

-   you want strings without typos or multiple versions of the same information

-   you want numeric variables to "make sense" (no infinity numbers, weird negative numbers etc.)

-   you want numbers as numeric variables

-   you want variables made of categories as factors
:::

In this lesson, we will explore how to modify existing columns and create new columns. The order of our column cleanings is as follows: 

1.   Cleaning character type variables and their strings for inconsistencies

2.   Cleaning data types to transform all categorical variables into factors

3.   With clean data types, we evaluate if our numerical variables are consistent

::: pro-tip
In this case it is important to have this order because we want to correct some of the character variables, and it is easier to do so while they are still characters. However, many of our numerical variable are factors so applying consistency checks of numerical variables to them would not make sense. So we convert categorical variables to factors before handling our numerical variables. 
:::

::: side-note
Much of the content covered in this lesson builds on what has already been covered in Chapter 4, Lesson 3 of the Introduction to Data analysis course. Have a look before starting !
:::

## Our data

::: recap
The data we are cleaning is the data from the COVID-19 serological survey conducted in Yaounde, Cameroon.

We will use the version of the dataset with:

-   standardized column names

-   no empty columns or rows

-   no duplicates.
:::

```{r, message = F, include=T, warning=FALSE}

#data frame
yaounde <- read.csv(here::here('ch02_data_cleaning_pipeline/data/yaounde_data_deduped.csv'))

```

## Dealing with inconsistencies

### Recode values {.unnumbered}

::: key-point
There are often times when you need to correct some inconsistencies in strings that might interfere with data analysis. This includes;

-   typos

-   capitalization errors.

-   misplaced punctuation

These issues can be fixed manually in the raw data source or we can make the change in the cleaning pipeline. The latter is more transparent and reproducible to anyone else seeking to understand or repeat your analysis.
:::

#### `recode()` {.unnumbered}

We can use the `recode()` function within the `mutate()` function to change specific values and to reconcile values not spelt the same.

:::vocab
The syntax is `recode(column_name, old_column_value = new_column_value)`
:::

We will have a look at the `age_category` column

```{r, message = F,warning=FALSE}
yaounde %>% count(age_category, name = "Count")
```

We see that there are 5 age groups with one of the groups being `May-14`. We will recode this category to `5-14`.

```{r, message = F,warning=FALSE}
yaounde %>%
  mutate(age_category = recode(age_category,`May-14` = "5-14")) %>%
  count(age_category, name = "Count")

```


We can also recode more than 1 value in a column. To do this, we will look at the variable `sex`

```{r, message = F,warning=FALSE}
yaounde %>% count(sex, name = "Count")
```

In this variable, there are capitalization errors and typos. We will recode values in the `sex` column so that `F = Female, female = Female,  M = Male and Mal = Male`

```{r, message = F,warning=FALSE}
yaounde %>%
  mutate(sex = recode(sex, `F` = "Female",
                         `female` = "Female",
                         `M` = "Male",
                         `Mal` = "Male")) %>%
  count(sex, name = "Count")
 
```

::: practice
1.   The variable `householdmembers` from the `typhoid` dataset should represent the number of individuals in a household. Display the different values in the variable. 

```{r,eval = FALSE,echo=FALSE}
q1 <- "YOUR ANSWER HERE"
.check_q1()
.hint_q1()
```

2. There is a value `01-May` in the `householdmembers` variable. Recode this value to `1-5`.
```{r,eval = FALSE,echo=FALSE}
q1 <- "YOUR ANSWER HERE"
.check_q1()
.hint_q1()
```

:::

#### `replace` {.unnumbered}

The `replace` function can be used to recode values with simple logical criteria . You can also use a logic condition to specify the rows to change. Within the replace function, we have to specify the name of our data object, the value you want to change and the replacement value.

::: vocab
The general syntax is:

`mutate(col_to_change = replace(col_to_change, criteria_for_rows, new value)`

OR 

`mutate(new_name = replace(col_to_change, criteria for rows, new value)`.

if you wan to create a new column
:::

One common situation to use `replace()` is changing just one value in one row, using an unique row identifier. This can be useful if you have external information indicating anomalies specific to a patient or to an observation and need to do a very specific edit. 

More generally, in cleaning, you will use `replace()` to change multiple rows based on some logic criteria. 

Conforming to human rights, we will indicate all patients under 18 years of age as having the `occupation` of `Child`. This is how we would do that: 


```{r, message = F,warning=FALSE}
yaounde %>% 
    mutate(occupation = replace(occupation, age < 18 , "Child"))
```

::: pro-tip
Use {dplyr's} `case_when()` if you are re-coding into many new groups, or if you need to use complex logic statements to re-code values.

This function evaluates every row in the data frame, assess whether the rows meets specified criteria, and assigns the correct new value.
:::

::: practice

3. The individual with `unique_key = 75` is described as "Housewhold head" in the variable `positioninthehousehold`. Replace this with "Household head"
```{r,eval = FALSE,echo=FALSE}
q3 <- "YOUR ANSWER HERE"
.check_q3()
.hint_q3()
```

:::

### String inconsistencies {.unnumbered}

#### `gsub` {.unnumbered} 

When fixing inconsistencies and structural issues in variables, it is often necessary to selectively replace multiple occurrences of a text within an R string.

The function `gsub()` is ideal for this as it searches for a regular expression in a string and replaces it.

The general syntax of the function is:

`gsub(search_term, replacement_term, string_searched)`

When combined with {dplyr's} `mutate()` function, a column of a data frame can be cleaned to enable analysis.

We will illustrate the use of this function using the `education` variable.

```{r, message = F,warning=FALSE}
yaounde %>% count(education, name = "Count")
```

There is an unwanted element in each of the categories ("1"). We will remove this element using `gsub` and replace it with an empty string "".

```{r, message = F,warning=FALSE}
yaounde %>%
    select(education)%>%
    mutate(education_new = gsub("1","",education))
```

Let us also change the encoding of `treatment_combinations` and `occupation` and replace all `--` with `/`

```{r, message = F,warning=FALSE}
yaounde %>%
    select(treatment_combinations,occupation)%>%
    mutate(across(c(treatment_combinations,occupation),
           ~ gsub("--","/",.x)))
```

::: pro-tip
Other useful checks in your column strings are:
-   finding whitespaces (`" "`) and replace them with (`"_"`) because whitespaces are never desirable in programming. You can reinsert them once you want to plot.
:::

::: practice
4.   Remove the value "1" from the variable `county`.

```{r,eval = FALSE,echo=FALSE}
q4 <- "YOUR ANSWER HERE"
.check_q4()
.hint_q4()
```

::: 

#### `str_replace_all` {.unnumbered} 

It is good practice to check for special characters (vector, known in R as `[[:punct:]]`) in your strings. This rigor is especially useful if you are working in another language than English (such as French), which uses accents, because accents are often transformed into special characters by computers. 

:::vocab
We will use `str_replace_all` from the {`stringr`} package, which allows you to input a vector of characters that you want to replace by another (commonly to remove them we use the `""` character).

The syntax is as follows:
`str_replace_all(column, vector_of_characters_to_replace, replacement_character)`
:::

```{r, message = F,warning=FALSE}
yaounde %>%
    mutate(across(where(is.character),
           ~ str_replace_all(.x, "[[:punct:]]","")))
```

#### Homogenize all strings throughout the dataset {.unnumbered}

Here, to easily manipulate all our data in the future, we transform all our strings to lowercase using the `tolower()` function. We select all character type columns with `where(is.character)`

```{r, message = F,warning=FALSE}
yaounde %>%
  mutate(across(where(is.character),
                ~ tolower(.x)))
```


::: practice
5.  Transform all the strings in the `typhoid` dataset to lowercase.

```{r,eval = FALSE,echo=FALSE}
q5 <- "YOUR ANSWER HERE"
.check_q5()
.hint_q5()
```

::: 

Now let's address data types. 

## Cleaning data types

Columns containing values that are numbers, factors or logical values (TRUE/FALSE) will only behave as expected if they are correctly classified. As such, you may need to redefine the type or class of your variable.

::: key-point
R has 6 basic data types/classes.

-   `character`: strings or individual characters, quoted
-   `numeric` : any real numbers (includes decimals)
-   `integer`: any integer(s)/whole numbers
-   `logical`: variables composed of `TRUE` or `FALSE`
-   `factor`: categorical/qualitative variables
-   `Date/POSIXct`: represents calendar dates and times

In addition to the ones listed above, there is also `raw` which will not be discussed in this lesson.
:::

::: recap 
In the introduction to this chapter we saw that our data has:
-   1 logical variable (`na` column which we removed because all values were `NA`)
-   20 character type variables 
-   15 numeric type variables 

Reminder, this is how we observed this:
```{r, eval=F, echo=T}
skim(yaounde) %>%
  select(skim_type) %>%
  count(skim_type)
```
:::

::: watch-out
There are no factor variables when you know the data contains variables such as age, sex: this means we will have quite some work in converting character or numeric variables to factors !
:::

::: key-point
**Why is converting categories to factors so important?** 

Factor variables are a way for representing categorical variables that can be either numeric or string variables. Factors are stored as integers, and often have labels associated with these unique integers. There are a number of advantages to converting categorical variables to factor variables;

-   Factor variables are also essential in rendering plots easily

-   Storing string variables as factor variables is a more efficient use of memory

-   Factor variables are essential when you want to apply a statistical model to your variable: the model will handle them correctly, assigning the correct number of degrees of freedom 

(if this does not ring any bells for you, no worries, we will delve into statistics into later courses: you will be happy to have your variables as factors then!)

Reminder: the functions to convert a variable to a factor are `as.factor()` from {dplyr} or `factor()` from base R 
:::

Now, let's move to cleaning our data types ! 

### Converting categories to factor type {.unnumbered}

Looking at our data, all our variables are categories, omit `age`, `height_cm`, `weight_kg`, and `respiratory_frequency`. 

::: side-note
Here, we already have `age_category` as a variable, else it would be important to create an age category variable: it is often useful to have age aggregates to look at data in plots or statistical analysis.
:::

```{r, message = F,warning=FALSE}
yaounde_factors <- 
  yaounde %>%
  mutate(across(!c(age, height_cm, weight_kg, respiration_frequency), 
                ~ as.factor(.x)))

skim(yaounde_factors) %>%
  select(skim_type) %>%
  count(skim_type)
```

::: side-note
As a reminder from your data wrangling lesson: we use `~` to indicate that we're supplying an anonymous function and use `.x` to indicate where the variables supplied in `across()` are used.
:::

::: practice

6. Convert the variables in position 13 to 29 to factor. 
```{r,eval = FALSE,echo=FALSE}
q6 <- "YOUR ANSWER HERE"
.check_q6()
.hint_q6()
```
:::

#### Using factors to recode values {.unnumbered}

All drug related variables are encoded as numeric variables. However, they are categories! They represent if a patient took (encoded by 1) or if they did not take (encoded by 0) that drug. 

::: pro-tip
If you want finer control while creating factors, use the `factor()` function. `as.factor()` should suffice in most cases but use `factor()` when you want to:

-   specify levels
-   modify labels
-   include NA as a level/category
-   create ordered factors
-   specify order of levels

The syntax is `factor(x,levels, labels = levels )`, where x is the variable in question.
:::

0 and 1 may not be very clear annotations of if a patient took a drug or not: let's transform all the drug variables into factors, while naming their levels with representative "No" (did not take the drug) instead of 0 and "Yes" (took the drug) instead of 1. 

```{r, message = F,warning=FALSE}
yaounde %>%
mutate(across(contains("drug"), # selects all variables reflecting a drug intake
              ~ factor(.x, # handles the variables one by one
                       levels = c(0:1), # selects the current binary varialbes
                       labels = c("No","Yes")))) %>% # and renames then with clearer annotations
  select(contains("drug")) 
```

::: pro-tip
For some manipulations, it could be better to have the encoding be binary (0 or 1), so this is an example data transformation, it's not imperative to do so.
:::

::: practice

7. Recode the variable `sex` to factor such that: `0 = female` and `1 = male`.

```{r,eval = FALSE,echo=FALSE}
q7 <- "YOUR ANSWER HERE"
.check_q7()
.hint_q7()
```
:::

## Back to inconsistencies

### Numeric inconsistencies {.unnumbered}

For numerical values, it is important to think about our variables and anticipate what errors would cause errors later on. 

A few examples: 
-   an abornmal occurence would be having negative values for variables that can only be positive (age, height, weight)
-   a problematic data recording includes have `Inf` values for numbers

Let's see how to address both of these issues. 

::: side-note
We will keep `NA` values and see how to address these more specifically in the next lesson 
:::

#### Ensuring non-negative variables {.unnumbered}

Here we will check that all our numeric variables are either `NA` (have not been entered), with the `is.na(.x)` condition, or are coherent i.e. are all positive values, with the `.x >= 0` condition. 

```{r, message = F,warning=FALSE, render=knit_print}
nrow(yaounde)
yaounde_positive_numeric <- 
  yaounde_factors %>%
  filter(across(where(is.numeric),
                ~ (.x >=0 | is.na(.x)) )) 

nrow(yaounde_positive_numeric)
```

From this filter, we see that none of our numeric variables had erroneous negative values: yay! It is always important to check this because you do not know until you check. 


### Ensuring finite variables {.unnumbered}

Here we will check that all our numeric variables are either `NA` (have not been entered), with the `is.na(.x)` condition, or are a finite number, with the `is.finite(.x)` condition. 

```{r, message = F,warning=FALSE, render=knit_print}
nrow(yaounde)
yaounde_finite_numeric <- 
  yaounde_factors %>%
  filter(across(where(is.numeric),
                ~ (is.finite(.x) | is.na(.x)) ))

nrow(yaounde_finite_numeric)
```

Once again, there are no strange values, such as `Inf`, in our dataset. Good that we checked. 

::: practice

8. Find the number of rows that are either `NA` or negative in the `typhoid` dataset.

```{r,eval = FALSE,echo=FALSE}
q8 <- "YOUR ANSWER HERE"
.check_q8()
.hint_q8()
```
:::

## Putting it all together 

Time to bring it all together, as you would to do a subsequent analysis. 

```{r, message = F,warning=FALSE, render=knit_print}
# Let's look with how many rows we started
nrow(yaounde)

yaounde_content_cleaned <- 
  yaounde %>%
  
  # STRING INCONSISTENCIES CORRECTIONS
  # Recode the age categories' strings
  mutate(age_category = recode(age_category,`May-14` = "5-14")) %>%
  # Recode the sex categories' strings
  mutate(sex = recode(sex, `F` = "Female",
                         `female` = "Female",
                         `M` = "Male",
                         `Mal` = "Male")) %>%
  # Change variable annotation for the occupation column
  mutate(occupation = replace(occupation, age < 18 , "Child")) %>%
  # Ensure in all columns that no strings are polluted by special characters
  mutate(across(where(is.character),
                ~ str_replace_all(.x, "[[:punct:]]",""))) %>%
  # Correct the education column's strings
  mutate(education = gsub("1","",education)) %>%
  # Homogenize all strings within all columns are similar: lowercase
  mutate(across(where(is.character),
                ~ tolower(.x))) %>%
  
  # CHANGE TYPE
  # Transform all categorical variables into factors
  mutate(across(!c(age, height_cm, weight_kg, respiration_frequency), 
              ~ as.factor(.x))) %>%
  # Make the drug treatment columns more readable to an external user
  mutate(across(contains("drug"), 
              ~ factor(.x, 
                       levels = c(0:1), 
                       labels = c("No","Yes")))) %>% 
    
  # NUMERIC INCONSISTENCIES CORRECTIONS
  # Check all numerical variables are positive (because in our case they should not be negative)
  filter(across(where(is.numeric), 
                         ~ (.x >=0 | is.na(.x)) )) %>%
  # Check all numerical variables are finite 
  filter(across(where(is.numeric),
                  ~ (is.finite(.x) | is.na(.x)) ))

# Let's see check we didn't loose any rows from cleaning !
# We are cleaning inside the rows so we expect to have THE SAME NUMBER of rows before and after
nrow(yaounde_content_cleaned)
```

Let's now save our cleaned dataset ! 

```{r, message = T,warning=FALSE}
write_csv(yaounde_content_cleaned,
          here::here('ch02_data_cleaning_pipeline/data/yaounde_content_cleaned.csv'))
```

## Wrapping up
In this chapter, we learnt how to fix inconsistencies and structural errors in the variables. However, there is still one more step in the data cleaning pipeline and that is; **dealing with missingness in the data.**
## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson: `r tgc_contributors_list(ids = c("eem", "lolovanco"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Horst, A. (2021). *Dplyr-learnr*. <https://github.com/allisonhorst/dplyr-learnr> (Original work published 2020)

Artwork was adapted from:

-   Horst, A. (2021). *R & stats illustrations by Allison Horst*. <https://github.com/allisonhorst/stats-illustrations> (Original work published 2018)

