---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: console
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)

# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

# Source autograder script quietly 
mute(here("ch04_functions/lessons/ch04_ls03_functions_autograder.R"))
```

# Functions: naming and commenting

## Learning objectives

1.  You will learn the conventions for naming functions.

2.  You will learn the importance of commenting your functions  (towards reproducible and easily understandable code!).

## General IMPORTANT advice when writing functions (part 1)

Before delving deeper into writing functions using conditional execution, there are a few points that should be said regarding functions.

### Naming {.unnumbered}

Always remember that functions communicate with the computer, but they also need to be comprehensible for humans to read (in the name of reproducible and collaborative programming). The name of a function should clearly state what it does while being short. Ideally, you want the name of your function to be short and informative, but this is not always possible, since it is a hard exercise. 

::: pro-tip
Remember always: **it’s better to be clear than short**. Remember that RStudio’s autocomplete functionality makes it easy to type long names quickly.
:::

As said by [Hadley Wickham](https://hadley.nz), **function names should be verbs, and arguments should be nouns**. 

::: side-note
Some exceptions to this are that nouns are ok if the function computes a very well known noun (i.e. `std_error()` is better than `compute_std_error()`). Also, it is ok if you are trying to access properties of an object (i.e. `coef()` is better than `get_coefficients()`). Use your best judgement and don’t be afraid to rename a function if you figure out a better name later.
:::

```{r eval=FALSE}

# Too short
f()

# Not a verb, or descriptive
a_great_function()

# Long, but clear
convert_utm()
age_months()

```

::: watch-out
Maintain consistency when naming your functions. This will be extremely useful if you manage to right your own R package in the future, and also if you share your functions with other colleagues. There aren't any rules written in stone for this. In the R universe, it is common to use **"snake_case" naming** (where each word is lower case and separated by an underscore), or **camelCase**.
:::

```{r eval=FALSE}

# Don't do mix naming styles: choose EITHER camelCase or snake_case
convert_utm()
ageMonths()

```

Also, try to have consistent names and arguments if you write functions that do similar things. A good practice is to use a common suffix in these cases, since RStudio`s autocomplete will show you other functions of the family

```{r eval=FALSE}

# Do this
convert_utm()
convert_wgs84()

# Don't do this
utm_convert()
wgs84_convert()

```

::: watch-out 
Avoid at all costs overriding existing functions and variables, especially from base R.

```{r eval=FALSE}

# Never ever do this
mean <- function(x) sum(x)
c <- function(x) print(x)

```
:::

::: practice

So far, what have you understood about naming your own functions in R? Type T (True) or F (False) to the sentences below: 

1. It's better for the function's name to be shorter than clear. 

2. It's recommended that function names should be verbs and arguments should be nouns. 

3. Try to adopt an standard when naming your functions, separating the words by `_` or using uppercase. For example, use `compute_utm()` or `computeUtm()`.

4. It's recommended to override existing functions in the base R as `mean()`.

:::

```{r include = F}

q1 <- "F"

.check_q1()

q2 <- "T"

.check_q2()

q3 <-"T"

.check_q3()

q4 <- "F"

.check_q4()

```

### Commenting {.unnumbered}

Finally, another good practice is to always comment your code. If you are not doing this already, make this a priority. One of the major advantages of commenting code is that you assure reproducibility not only for your future you but also other colleagues. Use and abuse of the `#` in your code to explain what you are doing and trying to achieve.

First, let's recap the expanded BMI function that we wrote in last lesson


```{r }

BMI <- function(weight, height, return_table = FALSE) { 
  
  x <- weight / (height ^ 2) 
  
  BMI_classes <- cut(x, 
                     breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf), 
                     labels = c("Underweight", "Normal weight", "Overweight", "Obesity class I", "Obesity class II", "Obesity class III"))
  
  BMI_table <- table(BMI_classes)
  
  if (return_table == TRUE) { 
    return(BMI_table)
  } else {
      return(x)
  }
}

```

Now, let's see an example of useful comments that will surely improve its documentation and help other users (and even your future you):

```{r eval=FALSE}

BMI <- function(weight, height, return_table = FALSE) { 
  # Calculating the BMI for each individual
  # weight should be in kilograms
  # height should be in meters
  x <- weight / (height ^ 2) 
  # creating categories from the BMI values
  BMI_classes <- cut(x, 
                     # creating breaks based on the accepted BMI thresholds
                     breaks = c(-Inf, 18.5, 24.9, 29.9, 34.9, 39.9, Inf), 
                     # using official classification to create categories
                     labels = c("Underweight", "Normal weight", "Overweight", "Obesity class I", "Obesity class II", "Obesity class III"))
  # creating a table that counts the observations in each BMI category
  BMI_table <- table(BMI_classes)
  # conditional statement; should the function return the BMI_table or the vector of numerical BMI values?
  if (return_table == TRUE) { 
    return(BMI_table)
  } else {
      return(x)
  }
}

```

This makes the function much more understandable, don't you agree? It is an exercise that you should be doing whenever writing functions and scripts in your career. 

```{r include = F}
.score_print()
```

## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("dcpcamara", "eduardo", "lolovanco", "kendavidn"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Barnier, Julien. "Introduction à R et au tidyverse." Accessed May 23, 2022. <https://juba.github.io/tidyverse>

-   Wickham, Hadley; Grolemund, Garrett. "R for Data Science." Accessed May 25, 2022. <https://r4ds.had.co.nz/>

<!-- (Chicago format. You can use https://www.citationmachine.net) -->

`r tgc_license()`
