---
title: '\ '
output: blogdown::html_page
---

```{r, echo = F, message = F, warning = F}
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, 
               here,
               knitr)

source(here("global/functions/misc_functions.R"))

opts_chunk$set(warning = F, message = F)
```

## Prologue

This quiz tests your understanding of the `filter()` function in the **dplyr** package. Good luck!

For multiple choice questions, please enter the correct letter option into the provided box.

------------------------------------------------------------------------


## Question 1

Which of the following is a **misinterpretation** of a **filter** statement on `df`? Answer: {a}

a.  `df %>% filter(deaths <  4)` : Drop row(s) where `deaths` is below 4

b.  `df %>% filter(!deaths >  4)` : Drop row(s) where `deaths` is above 4

c.  `df %>% filter(deaths ==  4)` : Keep row(s) where `deaths` is equal to 4

------------------------------------------------------------------------


## Question 2

<!--# Objective: Filter on NA-containing columns. -->

`df` is an excerpt of a edataset from an H7N9 outbreak in China.


```{r render = head_5_rows}
df <- outbreaks::fluH7N9_china_2013[c(1,2,3,74, 75), c(6:8)]
df
```

To **drop** the single female patient, which of the following **filter** statements is **correct** Answer: {a}

a.  `df %>% filter(gender ==  "m" & is.na(gender))`

b.  `df %>% filter(gender !=  "f")`

c.  `df %>% filter(gender ==  "m")`

------------------------------------------------------------------------


## Question 3

```{r echo = F}
df <- 
  tribble(~id, ~state, 
           1,  "Kano",
           2,  "Lagos",
           3,  "Bauchi", 
           4,  "Kano", 
           5,  "FCT",
           6,  NA)

df
```

From the dataset shown, you would like to keep rows where `state` is either "Bauchi" or "Kano". Which of the following uses **correct syntax** for this. Answer: {c}

a.  `df %>% filter(state %in% ("Bauchi" "Kano")`

b.  `df %>% filter(state %in% c("Bauchi" OR "Kano"))`

c.  `df %>% filter(state %in% c("Bauchi", "Kano"))`

------------------------------------------------------------------------


## Question 4

Answer T for TRUE or F for FALSE.

In R, `=` can be used as an assignment operator, while `==` is used for testing equality. Answer: {[T][TRUE]}

------------------------------------------------------------------------


