---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: console
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)

# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

# Source autograder script quietly 
here("ch99_template/lessons/ch99_ls02_practice_q_examples_autograder.R")
```

# Introduction

This document proposes some examples of autograders for the chapters on functions, conditionals and loops.

::: practice
-   Write a function that adds the number two to a vector.

```{r eval = F, include = F}
q1 <- function(){
}
.check_q1()
.hint_q1()
```

-   Let x equal the measure of the body temperature of patients that arrived at the emergency of your hospital. Write a function that returns the string "Fever" if the temperature, x, is equal to or above 38 degrees Celsius. (Note that the function does not need to work on more number at a time; doesn't need to work on a vector of length above 1.)

```{r eval = F, include = F}
q2 <- function(x){
  if(CONDITION_HERE) "OUTPUT_STRING_HERE"
}

.check_q2()
.hint_q2()
```

-   Using the `df2` data set below, create a loop that calculates the mean of each variable. Store it in a vector called `df2_means`, while defining its correct type and dimension.

```{r eval = F, include = F}
set.seed(1)
df2 <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rpois(10, 10),
  d = rnorm(10),
  e = rpois(10, 3),
  f = rnorm(10),
)

df2_means <- vector()
for (i in 1:ncol(df2)) {
  # Here, write code to populate the vector `df2_means` 
  # with the mean values of each column in `df2`
}
q3 <- df2_means

.check_q3()
```
:::



