---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: console
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)
library(dplyr)
# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

# Source autograder script quietly 
mute(here("ch05_loop_conditionals/lessons/ch05_ls04_loops_autograder.R"))
```

<!-- # Loops: automating tasks -->

## Learning objectives

1.  You will learn how to use `for` loops to automate tasks in R.

## Introduction to the dataset

In this lesson, we will use some data from the [Gapminder project](https://www.gapminder.org/). 
```{r render = reactable_10_rows, message = FALSE}

gapminder = read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
gapminder

```

## Automation with `for` loops

For loops can be really useful to automate tasks in R. 

Suppose we want to plot the variation in time of the life expectancy in Switzerland. We can do that using the code below. In the plot, we use as title: `Switzerland - lifeExp` and save the plot in the `images` folder.  

```{r} 
## filter the country to plot
gap_to_plot <- gapminder %>%
  filter(country == "Switzerland")

## plot
my_plot <- ggplot(data = gap_to_plot, aes(x = year, y = lifeExp)) + 
  geom_point() +
  ## add title and save
  labs(title = "Switzerland - lifeExp")

ggsave(filename = 'Switzerland_lifeExp.png', path = here("ch05_loop_conditionals/images"), plot = my_plot)
my_plot
```

To be able to create this plot with another country, let's replace the 'Switzerland' string by a variable called `ctry`. 

```{r} 

ctry = 'Switzerland'

gap_to_plot <- gapminder %>%
  filter(country == ctry)

## plot
my_plot <- ggplot(data = gap_to_plot, aes(x = year, y = lifeExp)) + 
  geom_point() +
  ## add title and save
  labs(title = paste(ctry, " - ", "lifeExp", sep = ""))

ggsave(filename = paste(ctry, '_lifeExp.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
my_plot
```

Suppose now, that we want to create this plot also for 'Brazil', 'Colombia' and 'Angola'. Instead of running the code above three times, we can create a loop that given a vector, change the value of `ctry` after each iteration. Let's do this in the code below: 

```{r} 

countries = c("Brazil", "Colombia", "Angola")

for (ctry in countries){
  gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes(x = year, y = lifeExp)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", "lifeExp", sep = ""))
  
  ggsave(filename = paste(ctry, '_lifeExp.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
  }

```

::: practice

 Based on the code above, select between the chunks of code below, which allows you to generate the plot for a list of countries and columns. The code must allow you to plot any column of the gapminder data set. 

* A
```{r eval = F, include = T}


countries = c("Brazil", "Colombia", "Angola")
cols_plot = c("pop", "lifeExp","gdpPercap")

for (col in countries){
  for (ctry in cols_plot) {
    
    gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes_string(x ="year", y = col)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", col, sep = ""))
  
  ggsave(filename = paste(ctry, '_' , col, '.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
  }
}

```

* B
```{r eval = F, include = T}


countries = c("Brazil", "Colombia", "Angola")
cols_plot = c("pop", "lifeExp","gdpPercap")

for (ctry in countries){
  for (col in cols_plot) {
    
    gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes_string(x ="year", y = lifeExp)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", col, sep = ""))
  
  ggsave(filename = paste(ctry, '_' , 'lifeExp', '.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
  }
}

```

* C
```{r eval = F, include = T}


countries = c("Brazil", "Colombia", "Angola")
cols_plot = c("pop", "lifeExp","gdpPercap")

for (ctry in countries){
  for (col in cols_plot) {
    
    gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes_string(x ="year", y = col)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", "lifeExp", sep = ""))
  
  ggsave(filename = paste(ctry, '_' , col, '.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
  }
}

```


* D
```{r eval = F, include = T}


countries = c("Brazil", "Colombia", "Angola")
cols_plot = c("pop", "lifeExp","gdpPercap")

for (ctry in countries){
  for (col in cols_plot) {
    
    gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes_string(x ="year", y = col)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", col, sep = ""))
  
  ggsave(filename = paste(ctry, '_' , col, '.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
  }
}

```
::: 

```{r, echo = FALSE}

q1 <- "D"

.check_q1()


```

We can also save the data filtered by country in a `.csv` file by adding the function `write.csv` in the loop.

```{r} 
countries = c("Brazil", "Colombia", "Angola")

for (ctry in countries){
  gap_to_plot <- gapminder %>%
    filter(country == ctry)
  
  ## plot
  my_plot <- ggplot(data = gap_to_plot, aes(x = year, y = lifeExp)) + 
    geom_point() +
    ## add title and save
    labs(title = paste(ctry, " - ", "lifeExp", sep = ""))
  
  setwd(here("ch05_loop_conditionals/data"))
  
  write.csv(gap_to_plot, paste("gapminder_",ctry, ".csv", sep = ""), row.names = FALSE)
  
  ggsave(filename = paste(ctry, '_lifeExp.png', sep = ""), path = here("ch05_loop_conditionals/images"), plot = my_plot)
}

 
```


```{r include = F}
.score_print()
```


--------------------------------------------------------------------------------

## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("dcpcamara", "eduardo", "lolovanco", "kendavidn"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   The Ocean Health Index Team. "Introduction to Open Data Science." Accessed Jun 10, 2022. <http://ohi-science.org/data-science-training/programming.html#automation-with-for-loops>


<!-- (Chicago format. You can use https://www.citationmachine.net) -->

`r tgc_license()`


