---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: console
---

```{r, include = FALSE, warning = FALSE, message = FALSE}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, knitr, here)

# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

# Source autograder script quietly 
mute(here("ch05_loop_conditionals/lessons/ch05_ls02_conditional_autograder.R"))
```

<!-- # Conditionals: advanced concepts -->

## Learning objectives

1.  You will learn more how to code complex chains of code using the `if/else if/ else` statement.

2.  You will learn the difference between `if/else`, `ifelse()` and `dplyr::if_else()`.

--------------------------------------------------------------------------------

## Conditional statement `if/else`, and the functions `ifelse` and `if_else`

In R, it is possible to use `if/else` or the base function `ifelse` (the `dplyr` package has the equivalent `if_else` function). Although both are very similar, the two are very different though.

- The `if/else` statements are used when you test a single condition and you want to execute different blocks of code depending on its result.

Suppose we have an object `x` that contains a **single** value, and we want to display a different message depending on whether it is less or greater than 30. In this case, we use `if/else`

```{r}

temperature <- 32

if (temperature >= 30) {
    message("Temperature is above 30 degrees Celsius")
} else {
    message("Temperature is below 30 degrees Celsius")
}

```

- The function `ifelse` applies a test to **all** elements of a **vector**. It returns a vector whose elements depend on the result of each test.

```{r}

temperatures <- runif(n = 10, min = 20, max = 42)

temperature_test <- ifelse(test = temperatures >= 30, 
                           yes = "Temperature is above 30 degrees Celsius", 
                           no = "Temperature is below 30 degrees Celsius")

temperature_test

```

The `if_else` function from the `dplyr` package is faster than the base `ifelse`. It is stricter also, because it checks if the `TRUE` and `FALSE` conditions are of the same type. It works nicely inside a `mutate` function (also, from the `dplyr` package). It also allows you to control missing values.

```{r render = reactable_10_rows}

library(dplyr)

temperatures <- runif(n = 10, min = 20, max = 42)

temperatures[4] <- NA

df_temperature <- as.data.frame(temperatures)

df_temperature %>% 
  mutate(temperature_test = if_else(condition = temperatures >= 30, 
                                    true = "Temperature is above 30 degrees Celsius", 
                                    false = "Temperature is below 30 degrees Celsius",
                                    missing = "Missing value!"))

```

You might be wondering what does "checking if the `TRUE` and `FALSE` conditions are of the same type" means. Check out this example. It returns an error since the `false` argument is not a character vector, but a numeric vector.

```{r}
df_temperature %>% 
  mutate(temperature_test = if_else(condition = temperatures >= 30, 
                                    true = "Temperature is above 30 degrees Celsius", 
                                    false = 999,
                                    missing = "Missing value!"))

```

::: pro-tip
If you need to categorize a number into a large number of factors, you might want to use the function `cut()` instead of a chain of `if/else if/else`. Read the help of `?cut` to understand it better. `cut` is a very useful function to categorize a numeric vector, such as a vector of BMI values into the categories established in the literature. 

- We passed the `x` vector as an argument of numeric vectors to the function.
- We passed a vector of cut points for the intervals as the argument called `breaks`.
- Finally, the argument `labels` received the names of the categories which will be created by the argument `breaks`.


```{r}

patient_temperatures <- runif(n = 10, min = 36.5, max = 43)

cut(x = patient_temperatures, 
    breaks = c(-Inf, 37.5, 38, 40, 42, Inf), 
    labels = c("No fever", "Fever", "High fever", "Immediate attention", "Allert"))

```

:::

::: practice
-   Let `x = runif(n = 10, min = 36.5, max = 42)` be a vector with body temperature measures of ten different patients that arrived at the emergency of your hospital. Use the `ifelse` or the `if_else` functions to write a test that identifies if a patient has fever when their temperature is equal or above 38 degrees Celsius. If the temperature is below 38, the patient does not have fever. The output messages should be: 'The patient has fever.', or 'The patient does not have fever.'.

::: watch-out
This time, since you are creating a vector of strings, *do not* use the function `message`.
:::

```{r, eval = F, echo = FALSE}
set.seed(1)
x = runif(n = 10, min = 36.5, max = 42)


q1 <- if_else(condition = x >= 38, 
              true = 'The patient has fever.', 
              false = 'The patient does not have fever.')

.check_q1()

.hint_q1()

```
:::

--------------------------------------------------------------------------------

## Advanced stuff regarding `if` conditions

This session draws heavily from the excellent advice from Hadley Wickham and Garrett Grolemund's "R for Data Science"<https://r4ds.had.co.nz/> book.

The `condition` in your chain must evaluate to either `TRUE` or `FALSE.` If it’s a vector or if it’s an `NA`, you’ll get an error.

```{r}

x <- c(1, 2, 1)

if (x) {}

if (NA) {}

```

Be very careful when using logical operators to test conditions. 

You should **never** use `|` (or) or `&` (and) in an if statement. These are vectorised operations that apply to multiple values, and that's why we use them inside the `filter()` function of the `dplyr` package. This means that they can be applied to vectors and will return a vector of `TRUE` and `FALSE`. Remember that `if`, `if/else` and `if/else if/else` are not applied over vectors!

::: reminder 
Inside `if`, `if/else` and `if/else if/else` statements, we should use `||` (or) and `&&` (and) as operators to combine multiple logical expressions. As soon as `||` sees the first `TRUE` it returns `TRUE` without computing anything else. As soon as `&&` sees the first `FALSE` it returns `FALSE`. This means that the double operators `||` and `&&` can only return a single value, and if given vectors they will only use the first value of each.

::: watch-out
Applying the & operator over a vector will return a vector of TRUE and FALSE values

```{r}

x <- 1:5

x > 0 & x <= 2

```
:::

::: watch-out
Applying the && operator over a vector will return just the first logical value (in this case, TRUE) and a warning message

```{r}

x <- 1:5

x > 0 && x <= 2
#> Warning messages:
#> 1: In x > 0 && x <= 2 : 'length(x) = 5 > 1' in coercion to 'logical(1)'

```
:::
::: 

When we pass a test to an `if` statement, it is supposed to return a single value `TRUE` or `FALSE`. A common mistake, especially when you are in a function (more on functions in a later lesson), is to apply an `if` statement to a vector. When this happens, R displays an error.

```{r}

x <- 1:5


if (x >= 5) {
    message("x is >= 5")
}

```

If you do have a logical vector, you can use `any()` or `all()` to collapse it to a single value.

```{r}

x <- c(TRUE, TRUE, FALSE, TRUE, FALSE)

```

::: watch-out
Wrong
```{r}

# wrong
if (x == TRUE) message("x is true")

```
:::

::: watch-out
Correct
```{r}

# correct
if (any(x) == TRUE) message("x is true") else message("x is false")
if (all(x) == TRUE) message("x is true") else message("x is false")

```
:::

Also, be careful when checking conditions for equality. The `==` operator is vectorised, which means that you can get a vector of outputs. You might have noticed that we used the `==` operator in almost all examples, but now you will understand better when to use it in a conditional statement:

To use the `==` operator, make sure to:

- Either check the `length` of the object is already 1.
- Collapse the object with `all` or `any`.
- Or use the non-vectorised `identical`. 

The function `identical` is very strict: it always returns either a single `TRUE` or a single `FALSE`. It also doesn’t coerce types. This means that you need to be careful when comparing integers and doubles:

```{r}

x <- c("Sarah", "Sahra", "Sara")

y <- "Sara"

any(x == y)

all(x == y)

identical(x, y)

```

Finally, make sure that when checking conditions involving numbers, you should be wary of floating point numbers. In these cases, you might want to use `dplyr::near` for comparisons.

```{r}

x <- sqrt(2) ^ 2

x

x == 2

```

Actually, `x` is not 2, as you just saw. It is a very, very small number:

```{r}

x - 2

```

In such cases, make sure to remember to use `dplyr::near` (if this is what you really want your condition to be):

```{r}

dplyr::near(x, 2)

```

```{r include = F}
.score_print()
```

--------------------------------------------------------------------------------

## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("dcpcamara", "eduardo", "lolovanco", "kendavidn"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Barnier, Julien. "Introduction à R et au tidyverse." Accessed May 23, 2022. <https://juba.github.io/tidyverse>

-   Wickham, Hadley; Grolemund, Garrett. "R for Data Science." Accessed May 25, 2022. <https://r4ds.had.co.nz/>

<!-- (Chicago format. You can use https://www.citationmachine.net) -->

`r tgc_license()`
