---
title: '\ '
output:
  html_document:
    number_sections: true
    toc: true
    css: !expr here::here("global/style/style.css")
    highlight: kate
    pandoc_args: --shift-heading-level-by=-1
  word_document:
    toc: true
editor_options:
  markdown:
    wrap: 100
  canonical: true
  chunk_output_type: inline
---

```{r, echo = F, message = F, warning = F}
# Load packages ----
if(!require(pacman)) install.packages("pacman")
pacman::p_load(clock,
               knitr,
               tidyverse, 
               here)

# Source functions 
source(here("global/functions/misc_functions.R"))

# knitr settings
knitr::opts_chunk$set(warning = F, message = F, class.source = "tgc-code-block", error = T)

## autograders
suppressMessages(source(here::here("ch01_dates_times/lessons/ch01_ls01_dates_times_intro_autograder.R")))
```

------------------------------------------------------------------------

<!-- # Date/time data: introduction -->

## Learning Objectives

1.  Learn about Build a date object by providing individual components (year, month, day) using `{clock}` and the `date_build()` function.
2.  Create a date-time object by providing individual components (year, month, day, hour, minute, second) using `{clock}` and the `date_time_build()` function.

## Overview

As a data analyst, being able to work with temporal data (that is data associated with dates and times) is a key skill to have.

Temporal data can be inputted in several ways.
For example, when recording the date a patient visited the clinic, one nurse could input the date this way - Monday, 3rd January, 2022 while another nurse could record the same date as 03/01/2022.
Thus, if our aim is to find out how many patients visited the hospital on a particular day, the data must be cleaned and the columns containing the date and/or time information must be reformatted.

### Date and time classes in R

Here are some fundamental things to know about working dates and times in R.

R has special classes to deal with date and/or time data.
We are going to discuss three of them.

-   The `date` class for dates (which can be formatted into different configurations of year-month-day.

-   The `time` class for times (hours, minutes and seconds)

-   The `date-time` class also called POSIXct used to represent date plus time that is to pin point a distinct moment in time on a particular day in specified time zone.

For example, to get the current date and time in your local time zone, R has two in-built functions.

`Sys.Date()` for the current date and `Sys.time()` for the current date-time in the time zone on your computer.

```{r now}
Sys.time()

Sys.Date()

```

## Creating date and times

Clock, is an R package that has many functions that make working with dates and times easier.

To install `{clock}`, run the following code in your R session.
`install.packages("clock")`.

Proceed to load the packages if you have already installed them.

### The `date_build()` function

We are going to use the function `date_build()` in the `{clock}` package to date objects with the birthdays given in the report above.

`clock::date_build()` is a function used to create a date by providing the following arguments.

-   The first argument provided is the year.
    If only one argument is given, the date output is the first day of the first month of that year.

    For example, run the code below, to quickly get the date- the first of January, 2022.

```{r year_date_build}
clock::date_build(2022)
```

You can specify the year, month and day by providing three integer values separated by commas.
For example, the output of the following code is 15th March, 2022.

```{r ides_of_March}
clock::date_build(2022,3,15)
```

::: watch-out
-   The second argument of the `clock::date_build()` function has a range of 1 to 12 to represent the 12 months (January to December).

-   The third argument of this function has a range of 1 to 31 to represent the number of possible days in a month.
:::

Thus, providing values outside these ranges would be invalid.
For example, the code below gives an error because there is no 14th month or 40th day of the year 2007.

```{r month_error}
try(clock::date_build(2007,14))
try(clock::date_build(2007,12,40))
```

You can create a range of dates using the colon `:` operator between the range you want either in the year, month or day arguments.

For example, to represent the first week of February 2002, in place of the third argument (day) in `clock::date_build(year, month, day)`,type `1:7` as shown below.

```{r day_error}
date_build(2002, 2, 1:7)
```

Also, to get the last day of each month the year 2015, you can type the following code in place of `clock::date_build(year,month, day)` .

-   The month argument is replaced by the range `1:12`

-   Then, the day argument is replaced by the string `"last"`

```{r last_day }
clock::date_build(2015,1:12,"last")
```

Supposing you are analyzing birth records from a hospital in Nairobi, Kenya in the report below.

> A healthy infant male was born on 27th December 2021 at 6 am.
> The next day, 3 baby girls were delivered.
> They were born in 1 hour intervals from 3pm.

::: practice
1.  Create date objects from this information and write code to output the baby boy's correct birthday.

    ```{r q-1, include = F}
    q1 <- "YOUR ANSWER HERE"
    .check_q1()
    .hint_q1a()
    .hint_q1b()
    ```
:::

### The `date_time_build()` function

To create a date-time object, a R object that contains both dates and times, we use the function `date_time_build()` with seven arguments `(year, month, day, hour, minute, second, zone)`.

A key thing, to note is how to input the arguments of date_time_build().
As discussed previously, month and day have specified ranges, so do hour, minute, second and zone.

-   The hour argument ranges from 0-23

-   The minute argument ranges from 0-59

-   The second argument ranges from 0-59

-   The zone is the local timezone written in the form "continent/city" eg.
    "America/New York" or "Africa/Lagos".
    You can check out the list of all time zones with associated cities by viewing this inbuilt R data set `OlsonNames()`

```{r list_timezones}
head(OlsonNames()) #Get first 6 rows of the timezones data set
```

For example, to create a date-time object of the boy born on 27th December, 2021 at 6 am in Nairobi, Kenya we should run the following code.

```{r date_time_build}
clock::date_time_build(2021,12,27,06,00,00, zone="Africa/Nairobi")
```

::: practice
2.  Create a data object for the birthdays of the three baby girls born the following day after 27th December, 2021 at 3pm with one hour interval between them.

    ```{r q-2,include = F}
    q2 <- "YOUR ANSWER HERE"
    .check_q2()
    .hint_q2a()

    .hint_q2b()
    .hint_q2c()

    .hint_q2d()
    .hint_q2e()

    .hint_q2f()

    ```
:::

```{r echo = F}
.score_print()
```

## Contributors {.unlisted .unnumbered}

The following team members contributed to this lesson:

`r tgc_contributors_list(ids = c("aowusuda"))`

## References {.unlisted .unnumbered}

Some material in this lesson was adapted from the following sources:

-   Wickham, Hadley, and Garrett Grolemund.
    "R For Data Science." 16 Dates and times \| R for Data Science, 2017.
    <https://r4ds.had.co.nz/dates-and-times.html#time-zones>.

-   Vaughan, Davis.
    "Building: Date-Time - date_time_build." - date_time_build • clock, February 12, 2021.
    <https://clock.r-lib.org/reference/date_time_build.html>.

`r tgc_license()`
